Finally Tagless Staged Interpreters for Simpler Typed Languages

### 2007-04-06 tagless/tagless-final.tex
related work is now ready to be edited

@@

We have been unable to find a direct reference that establishes that
the typed $\lambda$-calculus possesses a final co-algebra structure.
For the untyped $\lambda$-calculus, 
\citet{HonsellLenisa,honsell99coinductive} establish several interesting
results along this line.  Of particular interest is their use of
contexts with a hole \citep[p.13]{honsell99coinductive} for the definition
of \emph{observational equivalence}, which directly relates to our
use of terms with a hole as a context for our self-interpreter 
(\S\ref{selfinterp}).  The reader is directed to bibliography of
\citet{honsell99coinductive} for references to the foundational work
in this important area.

\jacques{Dig through the last couple of emails I just got from the
category theory list with references, and this section will be done.}

@@

\acks%\subsection*{Acknowledgments}
We thank Martin Sulzmann and Walid Taha 
for helpful discussions.

>>>>

% Allow things to flow a little more tightly around figures
\renewcommand\floatpagefraction{.95}
\renewcommand\topfraction{.95}
\renewcommand\bottomfraction{.95}
\renewcommand\textfraction{.05}

@@

We have been unable to find a direct reference that establishes that
the \emph{typed} $\lambda$-calculus possesses a final co-algebra structure.
For the untyped $\lambda$-calculus, 
\citet{HonsellLenisa,honsell99coinductive} establish several interesting
results along this line.  Of particular interest is their use of
contexts with a hole \citep[p.13]{honsell99coinductive} for the definition
of \emph{observational equivalence}, which directly relates to our
use of terms with a hole as a context for our self-interpreter 
(\S\ref{selfinterp}).  The reader is directed to bibliography of
\citet{honsell99coinductive} for references to the foundational work
in this important area.  Particularly intriguing is the link to the
coinductive aspects of B\"{o}hm trees, as pointed out by
\citet{berarducci-models} and \citet[Example 4.3.4]{jacobs-coalgebra}.

One way to interpret some of our results is that we eschew sums (at either the
term or type, or even kind level) for our terms, but instead use a record type 
(via Functors in OCaml and typeclasses in Haskell).
For the self-interpreter, we then proceed to use a Church encoding for
recursive data types \citep{bohm-automatic}.

@@

\acks%\subsection*{Acknowledgments}
We thank Martin Sulzmann and Walid Taha 
for helpful discussions.  Sam Staton, Pieter Hofstra, and Bart Jacobs
kindly provided some useful references on co-algebraic structures in the
$\lambda$\hyp calculus
